### 2015-03-09 doc/thesis/system.tex
Fixing typing rules in the appendix

As an example, lets assume that $x$ and $y$ are variables in the
environment with types $\Integer$ and $\String$, the assignment
\[
x, y = 1, ``foo"
\]
type checks through the following rules:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\dfrac{\env_{1}, \penv \vdash el:S_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash \overline{l}:S_{2}, \env_{3} \;\;\;
       S_{1} \lesssim S_{2}}
      {\env_{1}, \penv \vdash \overline{l} = el,\env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-EXPLIST1}\\
\dfrac{\env, \penv \vdash e_{i}:T_{i}, \env_{i} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\overline{e}\;|}
      {\env, \penv \vdash \overline{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{m}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env, \penv \vdash l_{i}:T_{i}, \env_{i} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\overline{l}\;|}
      {\env, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
\end{array}
\end{array}
\]

@@

The rule \textsc{T-CONSTRUCTOR2} type checks a table constructor with a non-empty
list of table fields:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\env_{i}, \penv \vdash p_{i}:(K_{i},V_{i}), \env_{i+1} \;\;\;
       n = |\;\vec{p}\;| \;\;\;
       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
       wf(T)}
      {\env_{1}, \penv \vdash \{\;\vec{p}\;\}:T, \env_{n+1}}
\end{array}
\]

@@

In this example, we assign the type $\{\}_{unique}$ to the variable
$person$, then we refine its type to $\{``firstname":\String\}_{unique}$,
and then we refine its type to $\{``firstname":\String, ``lastname":\String\}_{unique}$.
Rule \textsc{T-REFINE} type checks this use of refinement:
\[
\begin{array}{c}
\mylabel{T-REFINE}\\
\dfrac{\begin{array}{c}
       \env_{1}(n) = \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n} \}_{open|unique}\\
       \env_{1}, \penv \vdash k:K, \env_{2} \;\;\;
       \not \exists i \in 1..n \; K \lesssim K_{i} \;\;\;
       V = close(T)
       \end{array}}
      {\env_{1}, \penv \vdash n[k] {<}T{>}:V, \env_{2}[n \mapsto \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}, K{:}V\}_{open|unique}]}
\end{array}
\]

@@

Aliasing an \emph{unique} or an \emph{open} table type can produce
either a \emph{closed} or a \emph{fixed} table type, depending on
the context that we are using a variable.
Rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this behavior:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-IDREAD1}\\
\dfrac{\env(n) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env, \penv \vdash n:close(T_{2}), \env[n \mapsto open(T_{1})]}
\end{array}
\;
\begin{array}{c}
\mylabel{T-IDREAD2}\\
\dfrac{\env(n) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env, \penv \vdash n:fix(T_{2}), \env[n \mapsto fix(T_{1})]}
\end{array}
\end{array}
\]

@@

We also need to close \emph{unique} and \emph{open} tables that
appear in the left-hand side of assignments, as defined by
rule \textsc{T-IDWRITE}:
\[
\begin{array}{c}
\mylabel{T-IDWRITE}\\
\dfrac{\env(n) = T_{1} \;\;\; T_{2} = write(T_{1})}
      {\env, \penv \vdash n_{l}:close(T_{2}), \env[n \mapsto close(T_{2})]}
\end{array}
\]

@@

We also have different rules for type checking table indexing to avoid
changing table types in these operations, as they cannot create aliases:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-INDEX1}\\
\dfrac{\begin{array}{c}
       \env_{1}(n) = T \;\;\;
       read(\penv, T) = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}\\
       \env_{1}, \penv \vdash e_{2}:K, \env_{2} \;\;\;
       \exists i \in 1{..}n \; K \lesssim K_{i}
       \end{array}}
      {\env_{1}, \penv \vdash n[e_{2}]:V_{i}, \env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-INDEX2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash e_{1}:\{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}, \env_{2}\\
       \env_{2}, \penv \vdash e_{2}:K, \env_{3} \;\;\;
       \exists i \in 1{..}n \; K \lesssim K_{i}
       \end{array}}
      {\env_{1}, \penv \vdash e_{1}[e_{2}]:V_{i}, \env_{3}}
\end{array}
\end{array}
\]

A second form of refinement happens when we want to use an
\emph{unique} or \emph{open} table type in a context that expects a
\emph{fixed} or \emph{closed} table type with a different shape.
This kind of refinement allows programmers to add optional fields
or merge existing fields.
To do that, Typed Lua includes a type coercion expression that
obeys the \textsc{T-COERCE} rule below:
\[
\begin{array}{c}
\mylabel{T-COERCE}\\
\dfrac{\env_{1}(n) \subtype T \;\;\;
       \env_{1}[n \mapsto T], \penv \vdash n:T_{1}, \env_{2}}
      {\env_{1}, \penv \vdash {<}T{>} \; n:T_{1}, \env_{2}}
\end{array}
\]

@@

We also need to make sure to close all \emph{unique} and \emph{open}
table types before we type check a nested scope.
The rule \textsc{T-FUNCTION3} illustrates this case:
\[
\begin{array}{c}
\mylabel{T-FUNCTION3}\\
\dfrac{crall(\env_{1}[\vec{n} \mapsto \vec{T}]), \penv[\ret \mapsto S] \vdash s, \env_{2}}
      {\begin{array}{c}
       \env_{1}, \penv \vdash \mathbf{fun} \; (\vec{n{:}T}){:}S \; s:\vec{T} \times \Void \rightarrow S,\\
       openset(crset(\env_{1}, fav(\mathbf{fun} \; (\vec{n{:}T}){:}S \; s)),rv(\mathbf{fun} \; (\vec{n{:}T}){:}S \; s))
       \end{array}}
\end{array}
\]

Rule \textsc{T-FUNCTION3} uses some auxiliary functions to change
the type of variables before type checking a nested scope and
also to change the type of assigned and referenced variables after
type checking a nested scope.
The function \emph{crall} closes all \emph{unique} and \emph{open}
table types; it also restores filter types to their original type.
The function \emph{crset} closes a given set of free assigned variables,
which is given by the function \emph{fav}, and
it also uses this set to restore filter types to their original type.
The function \emph{openset} changes from \emph{unique} to \emph{open}
a given set of referenced variables, which is given by the function \emph{rv}.

@@

Our type system uses the pair $(x,S_{2})$ to allow rule
\textsc{T-LOCAL} to bound a union of second-level types $S_{2}$
to a variable $x$ in $\penv$:
\[
\begin{array}{c}
\mylabel{T-LOCAL}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash el:S_{1}, \env_{2}, (x,S_{2})\\
       \env_{2}[id_{1} \mapsto infer(S_{1},1), ..., id_{n} \mapsto infer(S_{1},n)], \penv[x \mapsto S_{2}] \vdash s, \env_{3} \;\;\;
       n = |\;\overline{id}\;|  
       \end{array}}
      {\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, \env_{3} - \{\overline{id}\}}
\end{array}
\]

@@

Introducing a variable $x$ in the projection environment allows our
type system to discriminate projection types $\pi_{i}^{x}$.
The rule \textsc{T-IF3} shows the case where our type system
discriminates a projection type based on the tag \texttt{nil}:
\[
\begin{array}{c}
\mylabel{T-IF3}\\
\dfrac{\begin{array}{c}
       \env_{1}(n) = \pi_{i}^{x} \;\;\; \penv(x) = S\\
       closeall(\env_{1}), \penv[x \mapsto fpt(S,\Nil,i)] \vdash s_{1}, \env_{2} \\
       closeall(\env_{1}), \penv[x \mapsto gpt(S,\Nil,i)] \vdash s_{2}, \env_{3} \\
       \env_{4} = openset(closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2})),rv(s_{1}) \cup rv(s_{2}))
      \end{array}}
      {\env_{1}, \penv \vdash \mathbf{if} \; n \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
\end{array}
\]

>>>>

As an example, lets assume that $x$ and $y$ are variables in the
environment with types $\Integer$ and $\String$, the assignment
\[
x, y = 1, ``foo"
\]
type checks through the following rules:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\dfrac{\env_{1}, \penv \vdash el:S_{1}, \env_{2} \;\;\;
       \env_{2}, \penv \vdash \overline{l}:S_{2}, \env_{3} \;\;\;
       S_{1} \lesssim S_{2}}
      {\env_{1}, \penv \vdash \overline{l} = el,\env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-EXPLIST1}\\
\dfrac{\env_{i}, \penv \vdash e_{i}:T_{i}, \env_{i+1} \;\;\;
       n = |\;\overline{e}\;| \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n+1})}
      {\env_{1}, \penv \vdash \overline{e}:T_{1} \times ... \times T_{n} \times \Nil{*}, \env_{m}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-LHSLIST}\\
\dfrac{\env, \penv \vdash l_{i}:T_{i}, \env_{i} \;\;\;
       \env_{m} = merge(\env_{1}, ..., \env_{n}) \;\;\;
       n = |\;\overline{l}\;|}
      {\env, \penv \vdash \overline{l}:T_{1} \times ... \times T_{n} \times \Value{*}, \env_{m}}
\end{array}
\end{array}
\]

@@

The rule \textsc{T-CONSTRUCTOR2} type checks a table constructor with a non-empty
list of table fields:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR2}\\
\dfrac{\env_{i}, \penv \vdash p_{i}:(K_{i},V_{i}), \env_{i+1} \;\;\;
       n = |\;\overline{p}\;| \;\;\;
       T = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}_{unique} \;\;\;
       wf(T)}
      {\env_{1}, \penv \vdash \{\;\overline{p}\;\}:T, \env_{n+1}}
\end{array}
\]

@@

In this example, we assign the type $\{\}_{unique}$ to the variable
$person$, then we refine its type to $\{``firstname":\String\}_{unique}$,
and then we refine its type to $\{``firstname":\String, ``lastname":\String\}_{unique}$.
Rule \textsc{T-REFINE} type checks this use of refinement:
\[
\begin{array}{c}
\mylabel{T-REFINE}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n} \}_{open|unique}\\
       \env_{1}, \penv \vdash k:K, \env_{2} \;\;\;
       \not \exists i \in 1..n \; K \lesssim K_{i} \;\;\;
       V = close(T)
       \end{array}}
      {\env_{1}, \penv \vdash id[k] {<}T{>}:V, \env_{2}[id \mapsto \{ K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}, K{:}V\}_{open|unique}]}
\end{array}
\]

@@

Aliasing an \emph{unique} or an \emph{open} table type can produce
either a \emph{closed} or a \emph{fixed} table type, depending on
the context that we are using a variable.
Rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this behavior:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-IDREAD1}\\
\dfrac{\env(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env, \penv \vdash id:close(T_{2}), \env[id \mapsto open(T_{1})]}
\end{array}
\;
\begin{array}{c}
\mylabel{T-IDREAD2}\\
\dfrac{\env(id) = T_{1} \;\;\; T_{2} = read(\penv, T_{1})}
      {\env, \penv \vdash id:fix(T_{2}), \env[id \mapsto fix(T_{1})]}
\end{array}
\end{array}
\]

@@

We also need to close \emph{unique} and \emph{open} tables that
appear in the left-hand side of assignments, as defined by
rule \textsc{T-IDWRITE}:
\[
\begin{array}{c}
\mylabel{T-IDWRITE}\\
\dfrac{\env(id) = T_{1} \;\;\; T_{2} = write(T_{1})}
      {\env, \penv \vdash id_{l}:close(T_{2}), \env[id \mapsto close(T_{2})]}
\end{array}
\]

@@

We also have different rules for type checking table indexing to avoid
changing table types in these operations, as they cannot create aliases:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-INDEX1}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = T \;\;\;
       read(\penv, T) = \{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}\\
       \env_{1}, \penv \vdash e_{2}:K, \env_{2} \;\;\;
       \exists i \in 1{..}n \; K \lesssim K_{i}
       \end{array}}
      {\env_{1}, \penv \vdash id[e_{2}]:V_{i}, \env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-INDEX2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash e_{1}:\{K_{1}{:}V_{1}, ..., K_{n}{:}V_{n}\}, \env_{2}\\
       \env_{2}, \penv \vdash e_{2}:K, \env_{3} \;\;\;
       \exists i \in 1{..}n \; K \lesssim K_{i}
       \end{array}}
      {\env_{1}, \penv \vdash e_{1}[e_{2}]:V_{i}, \env_{3}}
\end{array}
\end{array}
\]

A second form of refinement happens when we want to use an
\emph{unique} or \emph{open} table type in a context that expects a
\emph{fixed} or \emph{closed} table type with a different shape.
This kind of refinement allows programmers to add optional fields
or merge existing fields.
To do that, Typed Lua includes a type coercion expression that
obeys the \textsc{T-COERCE} rule below:
\[
\begin{array}{c}
\mylabel{T-COERCE}\\
\dfrac{\env_{1}(id) \subtype T \;\;\;
       \env_{1}[id \mapsto T], \penv \vdash id:T_{1}, \env_{2}}
      {\env_{1}, \penv \vdash {<}T{>} \; id:T_{1}, \env_{2}}
\end{array}
\]

@@

We also need to make sure to close all \emph{unique} and \emph{open}
table types before we type check a nested scope.
The rule \textsc{T-FUNCTION1} illustrates this case:
\[
\begin{array}{c}
\mylabel{T-FUNCTION1}\\
\dfrac{\begin{array}{c}
       crall(\env_{1}[\overline{id} \mapsto \overline{T}]), \penv[\ret \mapsto S] \vdash s, \env_{2}\\
       \env_{3} = openset(crset(\env_{1}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s)),rv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
       \end{array}}
      {\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}T}){:}S \; s:\overline{T} \times \Void \rightarrow S, \env_{3}}
\end{array}
\]

Rule \textsc{T-FUNCTION1} uses some auxiliary functions to change
the type of variables before type checking a nested scope and
also to change the type of assigned and referenced variables after
type checking a nested scope.
The function \emph{crall} closes all \emph{unique} and \emph{open}
table types; it also restores filter types to their original type.
The function \emph{crset} closes a given set of free assigned variables,
which is given by the function \emph{fav}, and
it also uses this set to restore filter types to their original type.
The function \emph{openset} changes from \emph{unique} to \emph{open}
a given set of referenced variables, which is given by the function \emph{rv}.

@@

Our type system uses the pair $(x,S_{2})$ to allow rule
\textsc{T-LOCAL2} to bound a union of second-level types $S_{2}$
to a variable $x$ in $\penv$:
\[
\begin{array}{c}
\mylabel{T-LOCAL2}\\
\dfrac{\begin{array}{c}
       \env_{1}, \penv \vdash el:S_{1}, \env_{2}, (x,S_{2})\\
       \env_{2}[id_{1} \mapsto infer(S_{1},1), ..., id_{n} \mapsto infer(S_{1},n)], \penv[x \mapsto S_{2}] \vdash s, \env_{3} \;\;\;
       n = |\;\overline{id}\;|  
       \end{array}}
      {\env_{1}, \penv \vdash \mathbf{local} \; \overline{id} = el \; \mathbf{in} \; s, \env_{3} - \{\overline{id}\}}
\end{array}
\]

@@

Introducing a variable $x$ in the projection environment allows our
type system to discriminate projection types $\pi_{i}^{x}$.
The rule \textsc{T-IF3} shows the case where our type system
discriminates a projection type based on the tag \texttt{nil}:
\[
\begin{array}{c}
\mylabel{T-IF3}\\
\dfrac{\begin{array}{c}
       \env_{1}(id) = \pi_{i}^{x} \;\;\; \penv(x) = S\\
       closeall(\env_{1}), \penv[x \mapsto fpt(S,\Nil,i)] \vdash s_{1}, \env_{2} \\
       closeall(\env_{1}), \penv[x \mapsto gpt(S,\Nil,i)] \vdash s_{2}, \env_{3} \\
       \env_{4} = openset(closeset(\env_{1}, fav(s_{1}) \cup fav(s_{2})),rv(s_{1}) \cup rv(s_{2}))
      \end{array}}
      {\env_{1}, \penv \vdash \mathbf{if} \; id \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}, \env_{4}}
\end{array}
\]
